---
title: "IRFinder"
author: "Florent Chuffart"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---



```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, results="hide")
```

# Method

1. We filter IRs according to IR change and adjusted p-value. Selected IRs are exported in .xlsx file named according to the condition and color.
2. We pool IRs that have been highlighted in at least one of the following volcano plots. Then we focus on this reduced list of IRs. 
3. We filter this reduced list of IRs according the correlation of their IR.change across conditions. Selected IRs are exported in .xlsx file named according to the condition and color.

```{r}
library(DESeq2)
source("/summer/epistorage/opt/IRFinder-1.3.0/bin/DESeq2Constructor.R")  #Load IRFinder-related function

# results = read.table("/summer/epistorage/opt/IRFinder-1.3.0/filePaths.txt")
# paths = as.vector(results$V1)                                            # File names must be saved in a vector
# experiment = read.table("/summer/epistorage/opt/IRFinder-1.3.0/experiment.txt",header=T)
# experiment$Condition=factor(experiment$Condition,levels=c("NHS","HS"))    # Set NHS as the baseline in the analysis
# rownames(experiment)=NULL                                                # Force removing rownames
#
# # WARNING: make sure the rownames of `experiment` is set to NULL.
# # WARNING: users MUST check if the order of files in the `path` matches the order of samples in `experiment` before continue
```


# Design

```{r design, results="verbatim"}
design = read.table("design.txt", header=TRUE, stringsAsFactors=FALSE); 
rownames(design) = design$SampleNames
design
```


```{r models}
da_questions = list(
  q1 = list(Lines=unique(design$Lines),         Condition=list(c("NHS", "HS"), c("NHS", "R3"))),
  q2 = list(Condition=unique(design$Condition), Lines=list(c("WT", "C1"), c("WT", "M3")))
)

if (!exists("res.diffs")) {
  res.diffs = list()
  for (nq in names(da_questions)) {
    q = da_questions[[nq]]
    for (var1 in unique(design[[names(q)[1]]])) {
      for (vars2 in q[[2]]) {
        idx_sample = design[design[[names(q)[1]]] %in% var1 & design[[names(q)[2]]] %in% vars2, ]$SampleNames
        print(design[idx_sample,])
    
        experiment = design[idx_sample,]
        paths = experiment$paths
        experiment = experiment[, c("SampleNames", names(q)[2])]
        rownames(experiment) = NULL
        experiment[[names(q)[2]]] = factor(experiment[[names(q)[2]]],levels=vars2)    # Set NHS as the baseline in the analysis
        if (!exists("mDESeqDataSetFromIRFinder")) {mDESeqDataSetFromIRFinder = memoise::memoise(DESeqDataSetFromIRFinder)}
        metaList = mDESeqDataSetFromIRFinder(filePaths=paths, designMatrix=experiment, designFormula=~1)
        # The above line generate a meta list contains four slots
        # First slot is a DESeq2 Object that can be directly pass to DESeq2 analysis.  
        # Second slot is a matrix for trimmed mean of intron depth
        # Third slot  is a matrix for correct splicing depth flanking introns
        # Fourth slot is a matrix for maximum splicing reads at either ends of introns
        # We build a “null” regression model on the interception only. 
        # A “real” model can be assigned either here directly, or in the downstream step. See below

        dds = metaList$DESeq2Object                       # Extract DESeq2 Object with normalization factors ready
        colData(dds)                                      # Check design of matrix

        # DataFrame with 8 rows and 3 columns
        #     SampleNames    Condition    IRFinder
        #     <factor>        <factor>    <factor>
        # 1   Sample1               WT          IR
        # 2   Sample2               WT          IR
        # 3   Sample3               KO          IR
        # 4   Sample4               KO          IR
        # 5   Sample1               WT          Splice
        # 6   Sample2               WT          Splice
        # 7   Sample3               KO          Splice
        # 8   Sample4               KO          Splice

        # Please note that sample size has been doubled and one additional column "IRFinder" has been added.
        # This is because IRFinder considers each sample has two sets of counts: one for reads inside intronic region and one for reads at splice site, indicating by "IR" and "Splice" respectively.
        # "IRFinder" is considered as an additional variable in the GLM model.
        # Please also be aware that size factors have been set to 1 for all samples. Re-estimation of size factors is NOT recommended and is going to bias the result.
        # More details at the end of the instruction.

        design(dds) = formula(paste0("~", names(q)[2], "+", names(q)[2], ":IRFinder"))     # Build a formula of GLM. Read below for more details. 
        if (!exists("mDESeq")) {mDESeq = memoise::memoise(DESeq)}
        dds = mDESeq(dds)                                  # Estimate parameters and fit to model
        resultsNames(dds)                                 # Check the actual variable name assigned by DESeq2

        # [1] "Intercept"                   "Condition_KO_vs_WT"
        # [3] "ConditionWT.IRFinderIR" "ConditionKO.IRFinderIR"

        res.ctrl = results(dds, name=paste0(names(q)[2], vars2[1], ".IRFinderIR"))
        # This tests if the number of IR reads are significantly different from normal spliced reads, in the NHS samples.
        # We might only be interested in the "log2FoldChange" column, instead of the significance.
        # This is because "log2FoldChange" represents log2(number of intronic reads/number of normal spliced reads).
        # So we the value of (intronic reads/normal spliced reads) by
    
        ctrl.IR_vs_Splice = 2^res.ctrl$log2FoldChange
    
        # As IR ratio is calculated as (intronic reads/(intronic reads+normal spliced reads))
        # We can easily convert the above value to IR ratio by
    
        IRratio.ctrl = ctrl.IR_vs_Splice/(1+ctrl.IR_vs_Splice)
    
        # Similarly, we can get IR ratio in the HS samples
        res.case = results(dds, name=paste0(names(q)[2], vars2[2], ".IRFinderIR"))
        case.IR_vs_Splice=2^res.case$log2FoldChange
        IRratio.case = case.IR_vs_Splice/(1+case.IR_vs_Splice)

        # Finally we can test the difference of (intronic reads/normal spliced reads) ratio between NHS and HS
        if (!exists("mDESeq2_results")) {mDESeq2_results = memoise::memoise(DESeq2::results)}
        res.diff = mDESeq2_results(dds, contrast=as.list(paste0(names(q)[2], vars2, ".IRFinderIR")))  
    
        # We can plot the changes of IR ratio with p values
        # In this example we defined significant IR changes as
        # 1) IR changes no less than 10% (both direction) and 
        # 2) with adjusted p values less than 0.05

        res.diff$IR.change = IRratio.case - IRratio.ctrl
    
        key = paste0("in_", var1, "_", paste(vars2, collapse="_ref_vs_"))
        res.diffs[[key]] = res.diff
      }
    }    
  }
}

```

# Volcano plots


```{r volcano}
layout(matrix(1:6, 2), respect=TRUE)

idx = c(
  "CLK1/ENSG00000013441/clean/2:200859746-200860124:-",
  "CLK1/ENSG00000013441/clean/2:200859747-200861237:-",
  "CLK1/ENSG00000013441/clean/2:200860215-200861237:-",
  "TAF1D/ENSG00000166012/clean/11:93738499-93739236:-",
  "DNAJB9/ENSG00000128590/clean/7:108570103-108571716:+",
  "DNAJB9/ENSG00000128590/clean/7:108571943-108572898:+"
)

intercepts_up = c(
  in_WT_NHS_ref_vs_HS=5,  
  in_WT_NHS_ref_vs_R3=4,  
  in_C1_NHS_ref_vs_HS=4,  
  in_C1_NHS_ref_vs_R3=4,    
  in_M3_NHS_ref_vs_HS=5,  
  in_M3_NHS_ref_vs_R3=5, 

  in_NHS_WT_ref_vs_C1=3,  
  in_NHS_WT_ref_vs_M3=3,  
   in_HS_WT_ref_vs_C1=2,  
   in_HS_WT_ref_vs_M3=3,    
   in_R3_WT_ref_vs_C1=4,  
   in_R3_WT_ref_vs_M3=3, 

  NULL 
)

intercepts_dw = c(
  in_WT_NHS_ref_vs_HS=4,  
  in_WT_NHS_ref_vs_R3=2,  
  in_C1_NHS_ref_vs_HS=2,  
  in_C1_NHS_ref_vs_R3=2,  
  in_M3_NHS_ref_vs_HS=3,  
  in_M3_NHS_ref_vs_R3=3, 

  in_NHS_WT_ref_vs_C1=1,  
  in_NHS_WT_ref_vs_M3=3,  
   in_HS_WT_ref_vs_C1=1,  
   in_HS_WT_ref_vs_M3=3,    
   in_R3_WT_ref_vs_C1=2,  
   in_R3_WT_ref_vs_M3=1, 


  NULL 
)







steep_updw = c(
  in_WT_NHS_ref_vs_HS=10,  
  in_WT_NHS_ref_vs_R3=10,  
  in_C1_NHS_ref_vs_HS=10,  
  in_C1_NHS_ref_vs_R3=10,    
  in_M3_NHS_ref_vs_HS=10,  
  in_M3_NHS_ref_vs_R3=10, 

  in_NHS_WT_ref_vs_C1=5,  
  in_NHS_WT_ref_vs_M3=5,  
   in_HS_WT_ref_vs_C1=5,  
   in_HS_WT_ref_vs_M3=5,    
   in_R3_WT_ref_vs_C1=5,  
   in_R3_WT_ref_vs_M3=5, 

  NULL 
)




myWriteXLS = WriteXLS::WriteXLS
# myWriteXLS = function(...){}
  
irs = list()
for (nq in names(da_questions)) {
  irs[[nq]] = list(red=list(), blue=list())
  q = da_questions[[nq]]
  for (var1 in unique(design[[names(q)[1]]])) {
    for (vars2 in q[[2]]) {
      key = paste0("in_", var1, "_", paste(vars2, collapse="_ref_vs_"))    
      print(key)
      res.diff = res.diffs[[key]]    
      tmpidx_up = rownames(res.diff)[
        !is.na(res.diff$padj) & !is.na(res.diff$IR.change) &
        res.diff$IR.change > 0 &
        -log10(res.diff$padj) > -steep_updw[[key]] *  res.diff$IR.change + intercepts_up[[key]] &
        TRUE
      ]
      tmpidx_dw = rownames(res.diff)[
        !is.na(res.diff$padj) & !is.na(res.diff$IR.change) &
        res.diff$IR.change < 0 &
        -log10(res.diff$padj) > steep_updw[[key]] *  res.diff$IR.change + intercepts_dw[[key]] &
        TRUE
      ]
      irs[[nq]][["red"]][[key]] = tmpidx_up
      irs[[nq]][["blue"]][[key]] = tmpidx_dw

      prefix = paste0(key, "_full")    
      tmpdf = as.data.frame(res.diff)
      head(tmpdf)
      myWriteXLS(tmpdf, paste0("volcano_", prefix, "_ir.xlsx"), row.names=TRUE, FreezeRow=1, FreezeCol=1)
      for (k in names(irs[[nq]])) {
        prefix = paste0(key, "_", k)    
        tmpdf = as.data.frame(res.diff[irs[[nq]][[k]][[key]],])
        head(tmpdf)
        myWriteXLS(tmpdf, paste0("volcano_", prefix, "_ir.xlsx"), row.names=TRUE, FreezeRow=1, FreezeCol=1)
        write.table(unique(do.call(rbind,strsplit(rownames(tmpdf), "/"))[,1]), paste0("volcano_", prefix, "_genes.txt"), col.names=FALSE, row.names=FALSE, quote=FALSE)
      }
    
      plot(
        # res.diff$IR.change, -log10(res.diff$pvalue),
        res.diff$IR.change, -log10(res.diff$padj), 
        main=paste(var1,"-",  paste(vars2, collapse=" (ref.) vs. "), length(tmpidx_up), length(tmpidx_dw)),
        col=ifelse(rownames(res.diff) %in% tmpidx_up, 2, ifelse(rownames(res.diff) %in% tmpidx_dw, 4, 1)),
        # col=adjustcolor("black", alpha.f=0.5),
        pch=".", 
        ylim=c(0,20)
      )
      text(res.diff[idx,]$IR.change, -log10(res.diff[idx,]$padj), 1:6, col=2) 
      legend("topleft", paste0("chr", substr(idx, 28, 1000)), pch=as.character(1:6), col=2,cex=.5, pt.cex=.5)
    }
  }
}

# sapply(irs[["red"]], length)
# length(unique(unlist(irs[["red"]])))
#
# sapply(irs[["blue"]], length)
# length(unique(unlist(irs[["blue"]])))
#
```

# Correlation plots

```{r correlation, fig.height=9, eval=TRUE}
# WT vs. C1 and WT vs. M3 in HS
# WT vs. C1 and WT vs. M3 in R3

layout(matrix(1:4, 2), respect=TRUE)

idx_clk1 = c(
  "CLK1/ENSG00000013441/clean/2:200859746-200860124:-",
  "CLK1/ENSG00000013441/clean/2:200859747-200861237:-",
  "CLK1/ENSG00000013441/clean/2:200860215-200861237:-",
  "TAF1D/ENSG00000166012/clean/11:93738499-93739236:-",
  "DNAJB9/ENSG00000128590/clean/7:108570103-108571716:+",
  "DNAJB9/ENSG00000128590/clean/7:108571943-108572898:+"
)


for (nq in names(da_questions)) {
  q = da_questions[[nq]]
  for (var1 in unique(design[[names(q)[1]]])[-1]) {
    for (vars2 in q[[2]]) {
      key1 = paste0("in_", unique(design[[names(q)[1]]])[1], "_", paste(vars2, collapse="_ref_vs_"))
      key2 = paste0("in_", var1                            , "_", paste(vars2, collapse="_ref_vs_"))
      print(paste0(key1, key2))
      res.diff1 = res.diffs[[key1]]    
      res.diff2 = res.diffs[[key2]]    
      idx  = unique(c(unlist(irs[[nq]][["red"]][1:2]), unlist(irs[[nq]][["blue"]][1:2])))
      idxs = list()
      idxs[["red"    ]] = intersect(idx, rownames(res.diff1)[res.diff1$IR.change > -res.diff2$IR.change & res.diff1$IR.change-.1 > res.diff2$IR.change])
      idxs[["blue"   ]] = intersect(idx, rownames(res.diff1)[res.diff1$IR.change > -res.diff2$IR.change & res.diff1$IR.change+.1 < res.diff2$IR.change])
      idxs[["violet" ]] = intersect(idx, rownames(res.diff1)[res.diff1$IR.change < -res.diff2$IR.change & res.diff1$IR.change+.1 < res.diff2$IR.change])
      idxs[["orange" ]] = intersect(idx, rownames(res.diff1)[res.diff1$IR.change < -res.diff2$IR.change & res.diff1$IR.change-.1 > res.diff2$IR.change])

      cols = rep("black", length(idx))
      names(cols) = idx
      for (k in names(idxs)) {
        cols[idxs[[k]]] = k
        prefix = paste0(k, "_", unique(design[[names(q)[1]]])[1], "vs", var1, "-",  paste(vars2, collapse="refvs"))    
        tmpdf = data.frame(idxs[[k]], res.diff1[idxs[[k]],]$IR.change, res.diff2[idxs[[k]],]$IR.change, stringsAsFactors=FALSE)
        colnames(tmpdf) = c("IR", paste0(c(key1, key2), "_IR.change"))
        head(tmpdf)
        myWriteXLS(tmpdf, paste0("correlation_", prefix, "_ir.xlsx"))
        write.table(unique(do.call(rbind,strsplit(tmpdf$IR, "/"))[,1]), paste0("correlation_", prefix, "_genes.txt"), col.names=FALSE, row.names=FALSE, quote=FALSE)
      }
    
      plot(res.diff1[idx,]$IR.change, res.diff2[idx,]$IR.change, 
        main=paste0(unique(design[[names(q)[1]]])[1], " vs. ", var1, " - ",  paste(vars2, collapse=" (ref.) vs. "), " (", length(idx), ")"),
        pch=".", 
        # col=adjustcolor("black", alpha.f=0.9),
        col=cols,
        xlim=c(-1,1), ylim=c(-1,1),
        xlab=paste0("IR.change ", unique(design[[names(q)[1]]])[1]), ylab=paste0("IR.change ", var1)
      )
    
      idx2 = idx[idx%in%names(cols)[cols!="black"]]
      points(res.diff1[idx2,]$IR.change, res.diff2[idx2,]$IR.change, 
        pch=1, 
        # col=adjustcolor("black", alpha.f=0.9),
        col=cols[idx2],
      )
    
      # plot(c(-1,1), c(-1,1))
      abline(a=0, b=1 , col="grey", lty=2)
      abline(a=0, b=-1, col="grey", lty=2)
      # text(res.diff1[intersect(idx_clk1, idx),]$IR.change, res.diff2[intersect(idx_clk1, idx),]$IR.change, 1:3, col=2)
      # legend("topleft", paste0("chr", substr(idx, 28, 1000)), pch=as.character(1:3), col=2)
      # abline(v=c(-.2,.2), h=c(-.2,.2), col="grey", lty=2)
      polygon(x=c( 1,  .05,    1),
              y=c(-1, -.05,  0.9), border="red"    ,col=adjustcolor("red"   , alpha.f=.1) , lty=2)
      polygon(x=c(-1, -.05,  0.9), 
              y=c( 1,  .05,    1), border="blue"   ,col=adjustcolor("blue"  , alpha.f=.1) , lty=2)
      polygon(x=c( 1,  .05, -0.9),
              y=c(-1, -.05, -  1), border="orange" ,col=adjustcolor("orange", alpha.f=.1) , lty=2)
      polygon(x=c(-1, -.05, -  1),
              y=c( 1,  .05, -0.9), border="violet" ,col=adjustcolor("violet", alpha.f=.1) , lty=2)
              
              
    }
  }
}

```
# Session Information

```{r, results="verbatim"}
sessionInfo()
```



